---
title: "Lab 9 - HPC"
output: 
html_document: default
link-citations: yes
---

# Learning goals

In this lab, you are expected to practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the parallel package.
- Use Rscript to submit jobs.

```{r eval=FALSE, echo=FALSE}
# install any missing packages
install.packages("microbenchmark")
```
```{r}
library(parallel)
```


## Problem 1

Give yourself a few minutes to think about what you learned about parallelization. List three
examples of problems that you believe may be solved using parallel computing,
and check for packages on the HPC CRAN task view that may be related to it.

_Answer here._
- Cross validation in machine learning. 
- caret -> supports parallel cross-validation with `doParallel`.
- mlr, foreach, doParallel -> for parallel model training

- bootstrapping
- boot -> bootstrapping
- parallel -> parallelize resampling

- markov chain monte carlo
- parallel
- rstan -> for stan for bayesian modeling
- RcppParallel -> parallel mcmc sampling
- nimle -> customize bayesian inference


## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without using
`parallel`:

1. This function generates a `n x k` dataset with all its entries having a Poisson distribution with mean `lambda`.

```{r p2-fun1}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  # YOUR CODE HERE
  matrix(rpois(n*k, lambda = lambda), ncol = k) # allow us to pre-allocating memory
  # rpois allows us to generate the random numbers all at once
}

# Benchmarking
microbenchmark::microbenchmark(
  fun1(100),
  fun1alt(100),
  unit = "ns"
)
```

How much faster?

_Answer here._
The fun1alt is about 10 times faster than the original fun1.

2.  Find the column max (hint: Checkout the function `max.col()`).

```{r p2-fun2}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max) # call max separately for each column
}

fun2alt <- function(x) {
  # YOUR CODE HERE
  x[cbind(max.col(t(x)), 1:ncol(x))] # avoids function calls inside loops
  # directly extracting the max values, without any loops
}

# Benchmarking
bench <- microbenchmark::microbenchmark(
  fun2(x),
  fun2alt(x),
  unit = "us"
)
bench
```

_Answer here with a plot._
```{r}
plot(bench)
ggplot2::autoplot(bench) +
  ggplot2::theme_bw()
```


## Problem 3: Parallelize everything

We will now turn our attention to non-parametric 
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.

The main assumption is that we can approximate many experiments by resampling
observations from our original dataset, which reflects the population. 

This function implements the non-parametric bootstrap:

```{r p3-boot-fun}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
 
  # Making the cluster using `ncpus`
  # STEP 1: GOES HERE
  cl <- makePSOCKcluster(ncpus) # creating a cluster for parallel computing
  # created worker nodes
  # ncpus specifying using multiple CPU cores
  # PSOCK is parallel socket cluster
  # R runs sequentially by default, cluster allows for parallel
  # STEP 2: GOES HERE
  # on.exit(stopCluster(cl)) # prevents memory leak
  # export the variables to the cluster
  clusterExport(cl, varlist = c("idx", "dat", "stat"), envir = environment()) # creating a new R session
  # sending the variables to all worker nodes
  # each run in isolated environment, don't have access to globel variable
  # idx -> resampling indices for bootstrapping
  # dat -> dataset
  # stat -> the statistical function that we use to compute the estimates
  
  # change sequential apply to parallelized apply
  # STEP 3: THIS FUNCTION NEEDS TO BE REPLACED WITH parLapply
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  # STEP 4: GOES HERE
  stopCluster(cl) # on.exit or stopCluster directly
  # to free up system resources
  
  ans
  
}
```

1. Use the previous pseudocode, and make it work with `parallel`. Here is just an example for you to try:

```{r p3-test-boot}
# Bootstrap of a linear regression model
my_stat <- function(d) coef(lm(y~x, data = d))

# DATA SIM
set.seed(1)
n <- 500 
R <- 1e4
x <- cbind(rnorm(n)) 
y <- x*5 + rnorm(n)

# Check if we get something similar as lm
# Ordinary least square CI
ans0 <- confint(lm(y~x))
cat("OLS CI \n")
print(ans0)

ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 4)
qs <- c(.025, .975)
cat("Bootstrp CI \n")
print(t(apply(ans1, 2, quantile, probs = qs)))


```

2. Check whether your version actually goes faster than the non-parallel version:

```{r benchmark-problem3}
# your code here
parallel::detectCores()

# non-parallel 1 core
system.time(my_boot(dat = data.frame(x, y), my_stat, R = 4000, ncpus = 1L))

# parallel 8 core
system.time(my_boot(dat = data.frame(x, y), my_stat, R = 4000, ncpus = 8L))
```

_Answer here._
user  system elapsed 
  0.029   0.008   1.023 
   user  system elapsed 
  0.107   0.047   0.974 

The first line should be the 8 cores result and the second line should be the 1 core result as 8 cores returned faster. Both user and system runtime are faster for parallelized process, but elapsed time is faster for 1 core process.

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command
in your terminal:

```bash
Rscript --vanilla -e 'rmarkdown::render("[full-path-to-your-Rmd-file.Rmd]")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full path to
your Rmd file... :).


